---
title: "Data Cleaning Procedures"
author: "Data Analysis Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

<style>
/* Applying Sagona font with fallbacks and increasing base font sizes */
body, p, li, td, th {
  font-family: "Sagona", "Georgia", "Times New Roman", serif;
  font-size: 18px;
  line-height: 1.6;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Sagona", "Georgia", "Times New Roman", serif;
  font-weight: bold;
  color: #2c3e50;
}

h1 { font-size: 32px; }
h2 { font-size: 26px; }
h3 { font-size: 22px; }

/* Styling code chunks and inline code for better color visibility */
code, pre {
  font-family: "Source Code Pro", "Courier New", monospace;
  font-size: 16px;
}

pre {
  background-color: #f4f4f4; /* Light gray background for code blocks */
  border: 1px solid #ddd;
  border-radius: 4px;
  padding: 10px;
  position: relative; /* Required for absolute positioning of the copy button */
}

p code, li code {
  background-color: #eaeaea;
  color: #c7254e;
  padding: 2px 4px;
  border-radius: 3px;
}

/* --- Copy Button Styling --- */
.copy-btn {
  position: absolute;
  top: 5px;
  right: 5px;
  background-color: #3498db;
  color: white;
  border: none;
  padding: 4px 8px;
  font-size: 12px;
  font-family: sans-serif;
  border-radius: 3px;
  cursor: pointer;
  opacity: 0; /* Hidden by default */
  transition: opacity 0.3s ease, background-color 0.3s ease;
}

pre:hover .copy-btn {
  opacity: 1; /* Show button when hovering over the code chunk */
}

.copy-btn:hover {
  background-color: #2980b9;
}

.copy-btn:active {
  background-color: #27ae60; /* Green flash when clicked */
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Loading required packages
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Introduction

Data cleaning is a crucial step in the data analysis pipeline. It ensures that the dataset is accurate, consistent, and reliable for further analysis or modeling. In this document, we will demonstrate three primary data cleaning procedures using the built-in R datasets `iris` and `mtcars`:

1. **Handling Missing Values**
2. **Removing Duplicates**
3. **Outlier Detection**

Since `iris` and `mtcars` are inherently clean datasets, we will artificially introduce missing values and duplicates to demonstrate the cleaning techniques effectively.

# 1. Handling Missing Values

Missing values can skew analysis and cause errors in statistical modeling. Common strategies include removing rows with missing data or imputing them with mean, median, or mode.

## Demonstration on `iris`

First, we will artificially introduce `NA` (Not Available) values into the `iris` dataset.

```{r}
# Introduce missing values in Sepal.Length and Petal.Width
iris_missing <- iris
iris_missing$Sepal.Length[c(5, 15, 30, 50, 34, 70, 130, 80, 144)] <- NA
iris_missing$Petal.Width[c(10, 40, 53, 81, 04, 52, 132)] <- NA

# Check for missing values
colSums(is.na(iris_missing))
```

**Visualizing Missing Data**

It is often helpful to visualize where missing data occurs. We can create a robust visualization by converting the data to a `TRUE`/`FALSE` matrix indicating missingness, and then plotting it.

```{r fig.width=8, fig.height=4}
# 1. Convert the dataset to a TRUE/FALSE matrix for missing values
missing_matrix <- as.data.frame(is.na(iris_missing))
missing_matrix$row_num <- 1:nrow(missing_matrix)

# 2. Pivot to long format for ggplot
missing_plot_df <- missing_matrix %>%
  pivot_longer(-row_num, names_to = "Variable", values_to = "Is_Missing")

# 3. Plot the missing data pattern
ggplot(missing_plot_df, aes(x = row_num, y = Variable, fill = Is_Missing)) +
  geom_raster() +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "lightgray"), name = "Missing?", labels = c("No", "Yes")) +
  labs(title = "Missing Data Pattern in Iris Dataset", x = "Row Index", y = "") +
  theme_minimal(base_size = 14)
```

**Strategy: Imputing NAs with the Mean**

We will impute the missing values with the column mean and visualize the distribution before and after to show the impact.

```{r fig.width=8, fig.height=4}
iris_imputed <- iris_missing %>%
  mutate(
    Sepal.Length = ifelse(is.na(Sepal.Length), mean(Sepal.Length, na.rm = TRUE), Sepal.Length),
    Petal.Width = ifelse(is.na(Petal.Width), mean(Petal.Width, na.rm = TRUE), Petal.Width)
  )

# Plotting distribution before and after imputation for Sepal.Length
df_vis <- data.frame(
  Value = c(iris_missing$Sepal.Length[!is.na(iris_missing$Sepal.Length)], iris_imputed$Sepal.Length),
  Status = c(
    rep("Original (NAs removed)", length(na.omit(iris_missing$Sepal.Length))),
    rep("After Imputation", nrow(iris_imputed))
  )
)

ggplot(df_vis, aes(x = Value, fill = Status)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 15) +
  labs(title = "Distribution of Sepal.Length: Original vs Imputed", x = "Sepal Length", y = "Count") +
  theme_minimal(base_size = 14)
```

# 2. Removing Duplicates

Duplicate rows can lead to over-representation of certain observations, biasing the results. We need to identify and remove them.

## Demonstration on `iris`

We will duplicate a few rows from the `iris` dataset and then use the `distinct()` function from the `dplyr` package to remove them.

```{r}
# Create duplicates by binding the first 5 rows to the original dataset
iris_duplicated <- rbind(iris, iris[1:5, ])

cat("Rows with duplicates:", nrow(iris_duplicated), "\n")

# Remove duplicates
iris_unique <- distinct(iris_duplicated)

cat("Rows after removing duplicates:", nrow(iris_unique))
```

**Visualizing Row Counts Before and After**

```{r fig.width=6, fig.height=4}
df_counts <- data.frame(
  Status = c("With Duplicates", "Duplicates Removed"),
  Count = c(nrow(iris_duplicated), nrow(iris_unique))
)

ggplot(df_counts, aes(x = Status, y = Count, fill = Status)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +
  labs(title = "Row Count Comparison: Iris Dataset", x = "", y = "Number of Rows") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# 3. Outlier Detection

Outliers are data points that differ significantly from other observations. They can be caused by measurement errors or natural variability. A common method for detecting outliers is the **Interquartile Range (IQR)** method. Data points falling below $Q1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$ are considered outliers.

## Demonstration on `iris`

Let's detect outliers in the `Sepal.Width` column of the `iris` dataset using boxplot statistics and visualize them on a scatter plot.

```{r fig.width=8, fig.height=5}
# Calculate Q1, Q3, and IQR
Q1 <- quantile(iris$Sepal.Width, 0.25)
Q3 <- quantile(iris$Sepal.Width, 0.75)
IQR_val <- Q3 - Q1

# Define upper and lower bounds
lower_bound <- Q1 - 1.5 * IQR_val
upper_bound <- Q3 + 1.5 * IQR_val

# Identify outliers and add a label
iris_outlier_vis <- iris %>%
  mutate(Outlier = ifelse(Sepal.Width < lower_bound | Sepal.Width > upper_bound, "Outlier", "Normal"))

# Visualizing outliers with a scatter plot (Sepal.Width vs Sepal.Length)
ggplot(iris_outlier_vis, aes(x = Sepal.Length, y = Sepal.Width, color = Outlier)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("Normal" = "steelblue", "Outlier" = "red")) +
  geom_hline(yintercept = c(lower_bound, upper_bound), linetype = "dashed", color = "gray50") +
  labs(
    title = "Outlier Detection in Iris (Sepal.Width)",
    subtitle = "Dashed lines represent IQR bounds (Q1 - 1.5*IQR, Q3 + 1.5*IQR)"
  ) +
  theme_minimal(base_size = 14)
```

## Demonstration on `mtcars`

We will apply the same IQR method to detect outliers in the `hp` (Horsepower) column of the `mtcars` dataset and visualize them using a bar chart.

```{r fig.width=8, fig.height=5}
# Calculate Q1, Q3, and IQR for hp
Q1_hp <- quantile(mtcars$hp, 0.25)
Q3_hp <- quantile(mtcars$hp, 0.75)
IQR_hp <- Q3_hp - Q1_hp

# Define bounds
lower_bound_hp <- Q1_hp - 1.5 * IQR_hp
upper_bound_hp <- Q3_hp + 1.5 * IQR_hp

# Identify outliers
mtcars_outlier_vis <- mtcars %>%
  mutate(
    Car = rownames(mtcars),
    Outlier = ifelse(hp < lower_bound_hp | hp > upper_bound_hp, "Outlier", "Normal")
  )

# Visualizing outliers with a bar chart
ggplot(mtcars_outlier_vis, aes(x = reorder(Car, hp), y = hp, fill = Outlier)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Normal" = "gray70", "Outlier" = "darkred")) +
  labs(
    title = "Outlier Detection in mtcars (Horsepower)",
    x = "Car Model",
    y = "Horsepower"
  ) +
  theme_minimal(base_size = 14)
```

# Conclusion

Through this document, we successfully demonstrated how to handle missing values via imputation and row removal, how to identify and filter out duplicate records, and how to detect outliers using the IQR method. By adding visualizations at each step, we can intuitively verify the effects of our data cleaning procedures, ensuring that datasets like `iris` and `mtcars` are prepped and ready for robust statistical analysis.

<script>
// JavaScript to add Copy Button to all code chunks
document.addEventListener("DOMContentLoaded", function() {
  const codeBlocks = document.querySelectorAll('pre');

  codeBlocks.forEach(function(block) {
    // Create the button element
    const button = document.createElement('button');
    button.className = 'copy-btn';
    button.textContent = 'Copy';

    // Add click event listener to the button
    button.addEventListener('click', function() {
      const code = block.querySelector('code');
      if (!code) return;

      const text = code.innerText;
      navigator.clipboard.writeText(text).then(() => {
        // Change button text to "Copied!" temporarily
        button.textContent = 'Copied!';
        button.style.backgroundColor = '#27ae60'; // Green flash

        setTimeout(() => {
          button.textContent = 'Copy';
          button.style.backgroundColor = '#3498db'; // Back to blue
        }, 1500);
      }).catch(err => {
        console.error('Failed to copy text: ', err);
      });
    });

    // Append the button to the code block
    block.appendChild(button);
  });
});
</script>
```
